---
title: 'Final Project: Beta-Dist Metropolis Hastings'
output: html_document
---
### M-H Code

```{r}
# Metropolis-Hastings function to run on a beta distribution with parameters (shape1, shape2) and proposal function with parameter c
mh.beta <- function(n.sims, burnin, shape1, shape2, c) {
  # Generates a random starting point between 0 and 1
  theta.cur <- runif(1,0,1)
  draws <- c()
  # Updating function that proposes a new value for the distribution and evaluates whether to add it to the list of accepted values
  theta.update <- function(theta.cur, shape1, shape2, c) {
    # Offers new candidate based on the given proposal function
    theta.can <- rbeta(1, c*theta.cur, c*(1-theta.cur))
    # Compares the pdf of the candidate value of that of the last value
    # If this ratio is high, there is a high chance the new value will be accepted
    accept.prob <- dbeta(theta.can, shape1 = shape1, shape2 = shape2)/dbeta(theta.cur,shape1 = shape1, shape2 = shape2)
    # Accept the proposed value into the distribution if a random value from 0 to 1 is less than it
    if (runif(1) <= accept.prob) theta.can else {theta.cur}
  }
  # Iterate the M-H algorithm n.sims times
  for (i in 1:n.sims) {
    draws[i] <- theta.cur <- theta.update(theta.cur, shape1 = shape1,
                                          shape2 = shape2, c=c)
  }
  return(draws[(burnin + 1):n.sims])
}

```

This Metropolis-Hastings algorithm allows for the proposal of beta distribution parameters and estimates values from the distribtion Beta~(shape1,shape2).  It uses the proposal function $$ \theta_{new} | \theta_{old} \sim Beta(c\theta_{old}, 1-c\theta_{old}) $$
to generate new proposal values based off the previously accepted one.  The algorithm then evaluates the new proposal beta pdf and compares it as a ratio to the old accepted beta pdf.  Finally, if this ratio is greater than a random number from 0 to 1, the propsed vaue is accepted.

```{r, echo = 1}
mh.draws <- mh.beta(10000, 0, 4, 6, 1)
trueBeta <- rbeta(10000, 4,6)

par(mfrow=c(2,2))  #2 rows, 2 columns
plot(mh.draws, main = 'Trace plot of draws'); acf(mh.draws, main = 'Autocorrelation plot of draws'); hist(mh.draws, main = 'Histogram of 10000 draws'); hist(trueBeta, main = 'Histogram of 10000 points on true beta(4,6) distribution')  #plot commands
```

Graphically, the M-H plot and the true Beta(4,6) plot look similar but notably different with n=10000, no burnIn, and c=1


```{r}
drawsCDF <- ecdf(mh.draws)
realCDF<- ecdf(trueBeta)
x <- seq(0,1,0.01)
max(drawsCDF(x)-realCDF(x))
```

A Kolmagorov-Smirnov test comparing the distributions under these conditions reveals a mild deviation from each other.

```{r}
mh.draws <- mh.beta(10000, 0, 4, 6, 0.1)

par(mfrow=c(2,2))  #2 rows, 2 columns
plot(mh.draws, main = 'Trace plot of draws'); acf(mh.draws, main = 'Autocorrelation plot of draws'); hist(mh.draws, main = 'Histogram of 10000 draws, c = 0.1'); hist(trueBeta, main = 'Histogram of 10000 points on true beta(4,6) distribution')  #plot commands

drawsCDF <- ecdf(mh.draws)
realCDF<- ecdf(trueBeta)
x <- seq(0,1,0.01)
max(drawsCDF(x)-realCDF(x))

```

```{r}
mh.draws <- mh.beta(10000, 0, 4, 6, 2.5)

par(mfrow=c(2,2))  #2 rows, 2 columns
plot(mh.draws, main = 'Trace plot of draws'); acf(mh.draws, main = 'Autocorrelation plot of draws'); hist(mh.draws, main = 'Histogram of 10000 draws, c = 2.5'); hist(trueBeta, main = 'Histogram of 10000 points on true beta(4,6) distribution')  #plot commands

drawsCDF <- ecdf(mh.draws)
realCDF<- ecdf(trueBeta)
x <- seq(0,1,0.01)
max(drawsCDF(x)-realCDF(x))

```

```{r}
mh.draws <- mh.beta(10000, 0, 4, 6, 10)

par(mfrow=c(2,2))  #2 rows, 2 columns
plot(mh.draws, main = 'Trace plot of draws'); acf(mh.draws, main = 'Autocorrelation plot of draws'); hist(mh.draws, main = 'Histogram of 10000 draws, c = 10'); hist(trueBeta, main = 'Histogram of 10000 points on true beta(4,6) distribution')  #plot commands

drawsCDF <- ecdf(mh.draws)
realCDF<- ecdf(trueBeta)
x <- seq(0,1,0.01)
max(drawsCDF(x)-realCDF(x))

```

Running the algorithm for different c values reveals that a higher c value produces a tighter correlation to the actual beta distribution.